Lattice Boltzmann simulations of the dynamics of thin liquid films \\ \large{Gitter Boltzmann Simulationen der Dynamik von d\"unnen Fl\"ussigkeitsfilmen}

### 2024-01-30 chapters/Method.tex
Printed and evaluated hopefully

\subsection{Finite differences}
Using finite differences means nothing more than discretizing the continuous problem into small (infinitesimal) discrete segments.
Figure~\ref{fig:finite_difference} illustrates a discretization approach where the function $f(x)$ is approximated by a finite set of points which are shown as black dots on the red line.
At every point $x_i$ along the x-axis we know the value of the function $f(x_i)$.
This however does not mean that we know the value of $f(x)$ at arbitrary $x$ values and even worse we do not know $\partial_x f(x_i)$ at $x_i$. 
For a continuous function, e.g., the film thickness $h(\mathbf{x},t)$, we can analytically compute the derivatives. 
Here we are left with a discrete set of points and have yet to come up with an idea to calculate $\nabla h(\mathbf{x}_i, t)$, which is crucial to compute the film pressure, Eq.~(\ref{eq:pfilm}).

@@

\subsection{Spectral methods}
Another strategy to solve a differential equation is to perform a transformation.
One rather famous transformation is the Fourier Transformation (FT), which we have used to find the dispersion relation in Sec.~\ref{sec:thin_films} and will encounter again in Chap.~\ref{chapter:second_paper}.
The set of solvers using a FT to solve a given differential equation is called spectral methods.
Doing so has the advantage of interchanging the derivatives with simple multiplications.
For example the term $\partial_x f(x)$ transforms to  
\begin{equation}\label{eq:fourier_transform}
    \partial_x f(x) = q\cdot \tilde{f}(q),
\end{equation}
where $\tilde{f}(q)$ is the Fourier transformed of $f(x)$, e.g., $\tilde{f}(q) = \delta(q - a/2\pi)$ for $f(x) = e^{iax}$. 
Instead of solving a differential equation one solves an algebraic equation, or a set of algebraic equations.
While it is often straightforward to perform the transformation it can become problematic to perform the backward transformation. 
Usually the expressions are fairly complicated and often the backward transformation into real space is not easy to perform.
In Chap.~\ref{chapter:second_paper} we arrive at such an expression for the structure factor $\mathcal{S}(q, t)$.
A structure factor is a measure of the distribution of length scales, roughly speaking.
In this specific case there is no need for a backward transformation because (scattering) experiments usually supply data in Fourier space.

@@

Figure~\ref{fig:scales_dummy} illustrates the idea coming from either a macroscopic system or a microscopic system.
Instead of dealing with every atom or molecule, which is displayed in the bottom left of the figure, it is also possible to use a more coarse approach which is shown in the centre of the figure where distributions over ensembles of particles are used~\cite{raabeOverviewLatticeBoltzmann2004}.
Therefore instead of working in the microscopic world the lattice Boltzmann method works in a mesoscopic regime.
Making it computationally more efficient than e.g., MD for a larger class of problems.
Of course, it is also possible to use a coarser approach and in fact a lot of the classical Navier-Stokes solvers operate in the upper right of Fig.~\ref{fig:scales_dummy}, see Sec.~\ref{sec:Joss_need} for some example solvers.

@@

So far however no assumptions on the lattice distributions are made and the discrete equilibrium distribution function has not been specified.
Paving work by He and Luo as well as group theoretical arguments derived by Rubinstein and Luo showed that based on dimensionality just a few (discrete) velocities are sufficient~\cite{heTheoryLatticeBoltzmann1997, rubinsteinTheoryLatticeBoltzmann2008}.
In fact in Chap.~\ref{chapter:first_paper} all numerical experiments were performed on a D2Q9 lattice therefore $\mathbf{c}_{i}$ is
\begin{equation}\label{eq:speeds_method}
\mathbf{c}_{i}  =
\left\{
\begin{array}{ll}
(0,0) & \alpha = 0, \\
\left[\cos{\frac{(i-1)\pi}{4}}, \sin{\frac{(i-1)\pi}{4}} \right] &  i=1,3,5,7 \\
\sqrt{2}\left[\cos{\frac{(i-1)\pi}{4}}, \sin{\frac{(i-1)\pi}{4}} \right] & i=2,4,6,8~,
\end{array}
\right.    
\end{equation}
The only missing pieces are $a, b, c, d$ of Eq.~(\ref{eq:eqi_dist1}).
Using the set of equations~(\ref{eq:moments_equilibria_1}-\ref{eq:moments_equilibria_4}) one quickly finds
\begin{equation}
    a = 1,\quad b = \frac{1}{c_s^2},\quad d = \frac{1}{2c_s^4},\quad e = \frac{1}{2c_s^2},
\end{equation}
with lattice weights
\begin{equation}\label{eq:weightsD2Q9_meth}
w_{i} = \begin{cases}
4/9 &\text{$i = 0$}\\
1/9 &\text{$i = 1,2,3,4$}\\
1/36 &\text{$i = 5,6,7,8$}
\end{cases}
,
\end{equation}
such that the equilibria can be written as
\begin{equation}\label{eq:eqi_dist2}
    f_{i}^{eq} = w_{i}\rho\left(1 + \frac{c_{i\alpha} u_{\alpha}}{c_s^2} +  \frac{(c_{i\alpha} u_{\alpha})^2}{2c_s^4} + \frac{u^2}{2c_s}\right).
\end{equation}
Together with the evolution equation Eq.~(\ref{eq:LBM_discret_noforces}) it is therefore possible to approximate the Navier-Stokes equation, at least in regimes of small Knudsen and Mach numbers, $Kn \ll 1$ and $Ma \ll 1$.

@@

In Chap.~\ref{chapter:first_paper} we will recap the main ideas of this derivation. 
We will then discuss constructive matching conditions between the shallow water model and the thin film equation.
The shallow water lattice Boltzmann model is not sufficient for thin film dynamics since we also have to introduce forces. 
These forces account for wetting dynamics, such as contact angles, and the velocity boundary condition at $h = 0$.

The derivation of the lattice Boltzmann method for the Navier-Stokes equation as well as the shallow water system concludes this chapter.
It follows a chapter about the sustainable approach I chose to keep the software first open source and maintainable for the future.
After the introduction of \textit{Swalbe.jl} in Chap.~\ref{chapter:fourth_paper} the method is thoroughly tested against problems such as the Rayleigh-Taylor instability in Chap.~\ref{chapter:first_paper}, while in Chap.~\ref{chapter:second_paper} an additional contribution due to thermal fluctuations is considered. 
Although the focus is solemnly on thermal fluctuations, this chapter should serve as a guideline for the inclusion of further dynamics to the thin film model.

Beyond additions to the thin film model we discuss a coupled problem in Chap.~\ref{chapter:third_paper}. 
Coupled in the sense that we not only consider the dynamics of the film but allow for dynamical degrees of freedom for the substrate.

>>>>

\subsection{Finite differences}
Using finite differences means nothing more than discretizing the continuous problem into small (infinitesimal) discrete segments.
Figure~\ref{fig:finite_difference} illustrates a discretization approach where the function $f(x)$ is approximated by a finite set of points which are shown as black dots on the red line.
At every point $x_i$ along the x-axis we know the value of the function $f(x_i)$.
This however does not mean that we know the value of $f(x)$ at arbitrary $x$ values and even worse we do not know $\partial_x f(x_i)$ at $x_i$. 
For a continuous function, e.g., the film thickness $h(\mathbf{x},t)$, we can analytically compute the derivatives. 
Here we are left with a discrete set of points and have yet to come up with an idea to calculate $\nabla h(\mathbf{x}_i, t)$, which is crucial to compute the film pressure.

@@

\subsection{Spectral methods}
Another strategy to solve a differential equation is to perform a transformation.
One rather famous transformation is the Fourier Transformation (FT), which we have used to find the dispersion relation in Sec.~\ref{sec:thin_films} and will encounter again in the third publication.
The set of solvers using a FT to solve a given differential equation is called spectral methods.
Doing so has the advantage of interchanging the derivatives with simple multiplications.
For example the term $\partial_x f(x)$ transforms to  
\begin{equation}\label{eq:fourier_transform}
    \partial_x f(x) = q\cdot \tilde{f}(q),
\end{equation}
where $\tilde{f}(q)$ is the Fourier transformed of $f(x)$, e.g., $\tilde{f}(q) = \delta(q - a/2\pi)$ for $f(x) = e^{iax}$. 
Instead of solving a differential equation one solves an algebraic equation, or a set of algebraic equations.
While it is often straightforward to perform the transformation it can become problematic to perform the backward transformation. 
Usually the expressions are fairly complicated and often the backward transformation into real space is not easy to perform.
In the third publication we arrive at such an expression for the structure factor $\mathcal{S}(q, t)$.
A structure factor is a measure of the distribution of length scales, roughly speaking.
In this specific case there is no need for a backward transformation because (scattering) experiments usually supply data in Fourier space.

@@

Figure~\ref{fig:scales_dummy} illustrates the idea coming from either a macroscopic system or a microscopic system.
Instead of dealing with every atom or molecule, which is displayed in the bottom left of the figure, it is also possible to use a more coarse approach which is shown in the centre of the figure where distributions over ensembles of particles are used~\cite{raabeOverviewLatticeBoltzmann2004}.
Therefore instead of working in the microscopic world the lattice Boltzmann method works in a mesoscopic regime.
Making it computationally more efficient than e.g., MD for a larger class of problems.
Of course, it is also possible to use a coarser approach and in fact a lot of the classical Navier-Stokes solvers operate in the upper right of Fig.~\ref{fig:scales_dummy}, see first publication for some example solvers.

@@

So far however no assumptions on the lattice distributions are made and the discrete equilibrium distribution function has not been specified.
Paving work by He and Luo as well as group theoretical arguments derived by Rubinstein and Luo showed that based on dimensionality just a few (discrete) velocities are sufficient~\cite{heTheoryLatticeBoltzmann1997, rubinsteinTheoryLatticeBoltzmann2008}.
In fact in second publication all numerical experiments were performed on a D2Q9 lattice therefore $\mathbf{c}_{i}$ is
\begin{equation}\label{eq:speeds_method}
\mathbf{c}_{i}  =
\left\{
\begin{array}{ll}
(0,0) & \alpha = 0, \\
\left[\cos{\frac{(i-1)\pi}{4}}, \sin{\frac{(i-1)\pi}{4}} \right] &  i=1,3,5,7 \\
\sqrt{2}\left[\cos{\frac{(i-1)\pi}{4}}, \sin{\frac{(i-1)\pi}{4}} \right] & i=2,4,6,8~,
\end{array}
\right.    
\end{equation}
The only missing pieces are $a, b, c, d$ of Eq.~(\ref{eq:eqi_dist1}).
Using the set of equations~(\ref{eq:moments_equilibria_1}-\ref{eq:moments_equilibria_4}) one quickly finds
\begin{equation}
    a = 1,\quad b = \frac{1}{c_s^2},\quad d = \frac{1}{2c_s^4},\quad e = \frac{1}{2c_s^2},
\end{equation}
with lattice weights
\begin{equation}\label{eq:weightsD2Q9_meth}
w_{i} = \begin{cases}
4/9 &\text{$i = 0$}\\
1/9 &\text{$i = 1,2,3,4$}\\
1/36 &\text{$i = 5,6,7,8$}
\end{cases}
,
\end{equation}
such that the equilibria can be written as
\begin{equation}\label{eq:eqi_dist2}
    f_{i}^{eq} = w_{i}\rho\left(1 + \frac{c_{i\alpha} u_{\alpha}}{c_s^2} +  \frac{(c_{i\alpha} u_{\alpha})^2}{2c_s^4} + \frac{u^2}{2c_s}\right).
\end{equation}
Together with the evolution equation Eq.~(\ref{eq:LBM_discret_noforces}) it is therefore possible to approximate the Navier-Stokes equation, at least in regimes of small Knudsen and Mach numbers, $Kn \ll 1$ and $Ma \ll 1$.

@@

In the second publication we will recap the main ideas of this derivation. 
We will then discuss constructive matching conditions between the shallow water model and the thin film equation.
The shallow water lattice Boltzmann model is not sufficient for thin film dynamics since we also have to introduce forces. 
These forces account for wetting dynamics, such as contact angles, and the velocity boundary condition at $h = 0$.

The derivation of the lattice Boltzmann method for the Navier-Stokes equation as well as the shallow water system concludes this chapter.
It follows a chapter about the sustainable approach I chose to keep the software first open source and maintainable for the future.
After the introduction of \textit{Swalbe.jl} in the first publication the method is thoroughly tested against problems such as the Rayleigh-Taylor instability in the second publication, while in the third publication an additional contribution due to thermal fluctuations is considered. 
Although the focus is solemnly on thermal fluctuations, this chapter should serve as a guideline for the inclusion of further dynamics to the thin film model.

Beyond additions to the thin film model we discuss a coupled problem in the last publication. 
Coupled in the sense that we not only consider the dynamics of the film but allow for dynamical degrees of freedom for the substrate.

### 2024-01-30 chapters/Theory.tex
Printed and evaluated hopefully

Mathematically the Navier-Stokes equation can be derived from one definition and two constraints.
The two constraints are mass and momentum conservation.
Mass conservation simply states that mass cannot be lost or created.
The same argument holds for the momentum.
By Noethers theorem every conserved quantity yields a symmetry and vice versa.
In this case conservation of momentum allows us to use the constraints on the solutions that translational symmetry requires.
Put simply a sliding drop experiment has to yield the same result in any laboratory on earths surface, independent of the exact location of the laboratory.
The last ingredient for the derivation is the definition of a fluid.
A fluid is ``\textit{a substance that flows and is not solid}`` according to the Cambridge Dictionary.
This definition can be refined, because a fluid is either a gas or a liquid that continuously deforms under shear stress, at least in the case of a Newtonian fluid.
And as such there is a linear relation between shear rate and shear velocity for a Newtonian fluid.
Clearly there are other liquids (which encloses granular systems to some extent) that exhibit different behaviour, for example toothpaste.
As long as there is no force applied to the tube the toothpaste does not flow.
Even the acceleration due to gravity when pointing the tube towards the ground does not exert enough force to generate a flow.
In contrast to Newtonian liquids those liquids have a different response function to shear, they can be addressed as \textit{power law} fluids.
While the phenomenology of these fluids, e.g., shear thinning or thickening, is very interesting and important in many industrial applications it is beyond the scope of this work. 
That said, in Chap.~\ref{chapter:second_paper} an extension towards more complex flows is presented. 
Instead of however changing the shear response we model thermal fluctuations to approximate the stochastic thin film equation (STF).
In fact, the step to derive an extension to our approach are nicely summarized in Chap.~\ref{chapter:second_paper}. 
It should serve as a guideline for further features, such as viscoelasticity for non-Newtonian liquids~\cite{bouchutNewModelShallow2013}.

@@

\section{Shallow water theory}
\label{sec:theory_shallow_water}
Solving the Navier-Stokes equation for the problems discussed in Chaps.~\ref{chapter:fourth_paper}-\ref{chapter:third_paper} is not needed.
Looking at Figs.~\ref{fig:examples_intro}-\ref{fig:morph_transition} one can identify that the height or thickness of the fluid layer is much smaller than the horizontal extension of the substrate.
This statement together with some more approximations allows to derive another system from the Navier-Stokes equation and the continuity equation which is called the shallow water or Saint-Venant equations~\cite{bTheorieMouvementNonpermanent1871}.

@@

Third, the horizontal velocity of the fluid has to vanish at the ``solid'' ground, or in terms of Fig.~\ref{fig:shallow_water_drawing} looking down from the free surface at $-h(x,y)$.
Although there is a viscosity associated with the fluid, friction with the bottom topography is the dominating dissipation channel.
This kind of boundary condition has been introduced in Chap.~\ref{chapter:intro} is called \textbf{no-slip} boundary condition~\cite{richardsonNoslipBoundaryCondition1973, degennesFluidWallSlippage2002, salmonLatticeBoltzmannMethod1999}.
While this may seem counterintuitive because fluid has no problem to flow over yet dry surfaces, modelling wetting or the wet-dry transition is a complex task, as seen later in this chapter and in Chap.~\ref{chapter:first_paper}.
After all, a theoretical model which is defined as a set of partial differential equations cannot account for the true complexity of nature~\cite{scrivenPhysicsApplicationsDIP1988}.
We come back to the idea of the no-slip boundary condition in Sec.~\ref{sec:thin_films}.

@@

To close this section we quickly collect the equations that form the shallow water system, see Eqs.~(\ref{eq:shallow_water_cont_true}, \ref{eq:sw_mom_final}).
The generalization of these equations towards a second horizontal dimension is straightforward and can be written as~\cite{salmonLatticeBoltzmannMethod1999, dellarNonhydrodynamicModesPriori2002, thommesLatticeBoltzmannMethods2007}
\begin{align}    
        \partial_t h &+ \nabla \cdot (h\mathbf{u}) = 0, \label{eq:shallow_water_01}\\
        \partial_t (h\mathbf{u}) + \nabla&\cdot (p \mathbb{1} + h\mathbf{u}\mathbf{u} - S) = 0, \label{eq:shallow_water_02} 
\end{align}
where $\mathbf{u} = (u_x, u_y)$ and the only unknown $\mathcal{S}$ is a yet undefined source term. 
The system Eqs.~(\ref{eq:shallow_water_01}-\ref{eq:shallow_water_02}) concludes the shallow water model. 
In Chap.~\ref{chapter:first_paper} we show how this translates into a numerical solver.

@@

In the shallow water section, we quickly reduced the problem to two dimensions (one horizontal and one vertical).
Here we keep the second horizontal dimension ($\tilde{y}, \tilde{u}_y$), because most numerical experiments in Chaps.~\ref{chapter:fourth_paper}-\ref{chapter:third_paper} are computed in a two dimensional domain.

@@

Now that the horizontal velocities are known we can perform the integration of the continuity equation Eq.~(\ref{eq:incomp})
\begin{equation}\label{eq:cont_thin}
    \int_0^h(\partial_x u_x + \partial_y u_y + \partial_z u_z) = 0,
\end{equation}
making use of the Leibniz rule one more time and showing only the $x$ component
\begin{equation}
    \int_0^h \partial_x u_x(z) \diff z = \partial_x\int_0^h u_x\diff z + \partial_x 0 u_x(0) - \partial_x h u_x(h) = \partial_x Q_x - \partial_x h u_x(h),  
\end{equation}
where Eq.~(\ref{eq:discharge_thin_x}) was used.
Inserting the above equation for the $x$ and $y$-component into the Eq.~(\ref{eq:cont_thin}) and performing the integration of $\partial_z u_z$ yields
\begin{equation}\label{eq:thin_dummy}
    \partial_x Q_x - (\partial_x h) u_x(h) + \partial_y Q_y - (\partial_y h) u_y(h) + u_z(h) - u_z(0) = 0.
\end{equation}
Since the no-slip boundary condition holds for all velocity components we set $u_z(0) = 0$.
The upper bound $u_z(h)$ can be computed using the material derivative, see Eq.~(\ref{eq:mat_dev}). 
We apply the same transformation as in Eq.~(\ref{eq:shallow_water_height}) with $f(\mathbf{x},t) = h(\mathbf{x},t) - z$, which is $0$ at the interface yields
\begin{equation}\label{eq:mat_uz_h}
    D_t f = \partial_t h + u_x\partial_x h + u_y\partial_y h - u_z = 0.
\end{equation}
Inserting the found result for $u(h)$ into Eq.~(\ref{eq:thin_dummy}) we get
\begin{equation}\label{eq:thin_obsc}
    \partial_x Q_x + \partial_y Q_y + \partial_t h = 0,
\end{equation}
which is almost the same as Eq.~(\ref{eq:simple_thin_film}).
Inserting $Q_x$ and $Q_y$, Eqs.~(\ref{eq:discharge_thin_x}) and (\ref{eq:discharge_thin_y}), we have
\begin{equation}
    \partial_t h(\mathbf{x},t) = \nabla\cdot\left(\frac{h^3}{3\mu}\nabla p + \frac{h^2}{2\mu}\tau\right),
\end{equation}
which in contrast to Eq.~(\ref{eq:simple_thin_film}) includes the shear contribution $\tau$.
Assuming the interface is not accelerated (e.g., in case of dewetting) the shear term can be neglected and the resulting equation is 
\begin{equation}\label{eq:thin_final}
    \partial_t h(\mathbf{x},t) = \nabla\cdot\left(\frac{h^3}{3\mu}\nabla p\right),
\end{equation}
which completes the task to derive the thin film equation from the Navier-Stokes equation. 
To sum up the approach in one sentence, we non dimensionalized the variables, collected terms in orders of $\varepsilon$, used our boundary conditions and performed an integration along the vertical dimension to average out the vertical dynamics.
Of course this derivation comes short if there are additional forces to consider such as thermal fluctuations, that said in Chap.~\ref{chapter:second_paper} we discuss the effect of thermal fluctuations on dewetting and show how the thin film equation has to be modified.

@@

An example of a dispersion relation which follows the above equation is shown in Fig.~\ref{fig:dispertion_1}.
In this figure we can identify two relevant wave numbers, on the one hand there is $q_0$ which is the fastest growing mode indicated by a dotted line and there is $q_m = \sqrt{2}q_0$ which is the largest wave number that is not damped.
Wavemodes larger than $q_m$ are therefore damped out.
A characteristic time scale of the problem is given as
\begin{equation}
    t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4},
\end{equation}
with is used to normalize the time in Fig.~\ref{fig:dispertion_1}.
Because the concept of linear stability is such an important one, we encounter it again in Chap.~\ref{chapter:first_paper} (see Fig.~\ref{fig:RTI}) as well as in Chap.~\ref{chapter:second_paper} (see Figs.~\ref{fig:theory_simulation_structure_factor}, \ref{fig:spectral_analysis_deter_sine8}, \ref{fig:spectral_analysis_stoch_sine8}, \ref{fig:square_wave_both}).
Chap.~\ref{chapter:second_paper} focuses only on the addition of a fluctuating term and the so called stochastic thin film equation.
The new term due to thermal fluctuations offers some interesting features on both the structure factor as well as the dispersion.

\begin{figure}
    \centering
    \includegraphics[width=0.65\textwidth]{graphics/spectratheta20.pdf}
    \caption{Dispersion relation of a spinodally dewetting thin film with and without thermal fluctuations.
    Where $S(q)$ is the structure factor depending on the wavemode $q$.
    The dots and triangles are generated using the here presented numerical method. 
    Full and dashed lines are the theoretical prediction based on Eq.~(\ref{eq:simple_thin_film}) and its fluctuating extension, Eq.~(\ref{eq:structure_factor_2}).
    }
    \label{fig:dispertion_1}
\end{figure}

@@

While there are many similarities between the thin film equation and the shallow water equations, there are also some major differences.
The most pressing difference between these two models is the fact that the shallow water equation is a vector equation while the thin film equation is a scalar equation.
In the shallow water equations the liquid/vapour interface is driven by hydrostatic pressure while in the thin film equation we neglect the hydrostatic pressure and introduce surface tension and capillarity.
The question that naturally arise here is: can one model be translated into the other one?
Although this question seems odd at first it is actually of great significance to show that problems can be solved with different approaches.
There are several numerical methods to iteratively solve the thin film equation.
There are as well several different numerical methods to approximate the Navier-Stokes equation.
The important contribution to our understanding of CFD is that methods build to solve the shallow water equations can actually be used to solve the thin film problem, as discussed in Chap.~\ref{chapter:first_paper}

>>>>

Mathematically the Navier-Stokes equation can be derived from one definition and two constraints.
The two constraints are mass and momentum conservation.
Mass conservation simply states that mass cannot be lost or created.
The same argument holds for the momentum.
By Noethers theorem every conserved quantity yields a symmetry and vice versa.
In this case conservation of momentum allows us to use the constraints on the solutions that translational symmetry requires.
Put simply a sliding drop experiment has to yield the same result in any laboratory on earths surface, independent of the exact location of the laboratory.
The last ingredient for the derivation is the definition of a fluid.
A fluid is ``\textit{a substance that flows and is not solid}`` according to the Cambridge Dictionary.
This definition can be refined, because a fluid is either a gas or a liquid that continuously deforms under shear stress, at least in the case of a Newtonian fluid.
And as such there is a linear relation between shear rate and shear velocity for a Newtonian fluid.
Clearly there are other liquids (which encloses granular systems to some extent) that exhibit different behaviour, for example toothpaste.
As long as there is no force applied to the tube the toothpaste does not flow.
Even the acceleration due to gravity when pointing the tube towards the ground does not exert enough force to generate a flow.
In contrast to Newtonian liquids those liquids have a different response function to shear, they can be addressed as \textit{power law} fluids.
While the phenomenology of these fluids, e.g., shear thinning or thickening, is very interesting and important in many industrial applications it is beyond the scope of this work. 
That said, in the third publication an extension towards more complex flows is presented. 
Instead of however changing the shear response we model thermal fluctuations to approximate the stochastic thin film equation (STF).
In fact, the step to derive an extension to our approach are nicely summarized in the third publication. 
It should serve as a guideline for further features, such as viscoelasticity for non-Newtonian liquids~\cite{bouchutNewModelShallow2013}.

@@

\section{Shallow water theory}
\label{sec:theory_shallow_water}
Solving the Navier-Stokes equation for the problems discussed in all publications is not needed.
Looking at Figs.~\ref{fig:examples_intro}-\ref{fig:morph_transition} one can identify that the height or thickness of the fluid layer is much smaller than the horizontal extension of the substrate.
This statement together with some more approximations allows to derive another system from the Navier-Stokes equation and the continuity equation which is called the shallow water or Saint-Venant equations~\cite{bTheorieMouvementNonpermanent1871}.

@@

Third, the horizontal velocity of the fluid has to vanish at the ``solid'' ground, or in terms of Fig.~\ref{fig:shallow_water_drawing} looking down from the free surface at $-h(x,y)$.
Although there is a viscosity associated with the fluid, friction with the bottom topography is the dominating dissipation channel.
This kind of boundary condition has been introduced in Chap.~\ref{chapter:intro} is called \textbf{no-slip} boundary condition~\cite{richardsonNoslipBoundaryCondition1973, degennesFluidWallSlippage2002, salmonLatticeBoltzmannMethod1999}.
While this may seem counterintuitive because fluid has no problem to flow over yet dry surfaces, modelling wetting or the wet-dry transition is a complex task, as seen later in this chapter and in the second publication.
After all, a theoretical model which is defined as a set of partial differential equations cannot account for the true complexity of nature~\cite{scrivenPhysicsApplicationsDIP1988}.
We come back to the idea of the no-slip boundary condition in Sec.~\ref{sec:thin_films}.

@@

To close this section we quickly collect the equations that form the shallow water system, see Eqs.~(\ref{eq:shallow_water_cont_true}, \ref{eq:sw_mom_final}).
The generalization of these equations towards a second horizontal dimension is straightforward and can be written as~\cite{salmonLatticeBoltzmannMethod1999, dellarNonhydrodynamicModesPriori2002, thommesLatticeBoltzmannMethods2007}
\begin{align}    
        \partial_t h &+ \nabla \cdot (h\mathbf{u}) = 0, \label{eq:shallow_water_01}\\
        \partial_t (h\mathbf{u}) + \nabla&\cdot (p \mathbb{1} + h\mathbf{u}\mathbf{u} - S) = 0, \label{eq:shallow_water_02} 
\end{align}
where $\mathbf{u} = (u_x, u_y)$ and the only unknown $\mathcal{S}$ is a yet undefined source term. 
The system Eqs.~(\ref{eq:shallow_water_01}-\ref{eq:shallow_water_02}) concludes the shallow water model. 
In second publication we show how this translates into a numerical solver.

@@

In the shallow water section, we quickly reduced the problem to two dimensions (one horizontal and one vertical).
Here we keep the second horizontal dimension ($\tilde{y}, \tilde{u}_y$), because most numerical experiments in the last and second to last publication are computed in a two-dimensional domain.

@@

Now that the horizontal velocities are known we can perform the integration of the continuity equation Eq.~(\ref{eq:incomp})
\begin{equation}\label{eq:cont_thin}
    \int_0^h(\partial_x u_x + \partial_y u_y + \partial_z u_z) = 0,
\end{equation}
making use of the Leibniz rule one more time and showing only the $x$ component
\begin{equation}
    \int_0^h \partial_x u_x(z) \diff z = \partial_x\int_0^h u_x\diff z + \partial_x 0 u_x(0) - \partial_x h u_x(h) = \partial_x Q_x - \partial_x h u_x(h),  
\end{equation}
where Eq.~(\ref{eq:discharge_thin_x}) was used.
Inserting the above equation for the $x$ and $y$-component into the Eq.~(\ref{eq:cont_thin}) and performing the integration of $\partial_z u_z$ yields
\begin{equation}\label{eq:thin_dummy}
    \partial_x Q_x - (\partial_x h) u_x(h) + \partial_y Q_y - (\partial_y h) u_y(h) + u_z(h) - u_z(0) = 0.
\end{equation}
Since the no-slip boundary condition holds for all velocity components we set $u_z(0) = 0$.
The upper bound $u_z(h)$ can be computed using the material derivative, see Eq.~(\ref{eq:mat_dev}). 
We apply the same transformation as in Eq.~(\ref{eq:shallow_water_height}) with $f(\mathbf{x},t) = h(\mathbf{x},t) - z$, which is $0$ at the interface yields
\begin{equation}\label{eq:mat_uz_h}
    D_t f = \partial_t h + u_x\partial_x h + u_y\partial_y h - u_z = 0.
\end{equation}
Inserting the found result for $u(h)$ into Eq.~(\ref{eq:thin_dummy}) we get
\begin{equation}\label{eq:thin_obsc}
    \partial_x Q_x + \partial_y Q_y + \partial_t h = 0,
\end{equation}
which is almost the same as Eq.~(\ref{eq:simple_thin_film}).
Inserting $Q_x$ and $Q_y$, Eqs.~(\ref{eq:discharge_thin_x}) and (\ref{eq:discharge_thin_y}), we have
\begin{equation}
    \partial_t h(\mathbf{x},t) = \nabla\cdot\left(\frac{h^3}{3\mu}\nabla p + \frac{h^2}{2\mu}\tau\right),
\end{equation}
which in contrast to Eq.~(\ref{eq:simple_thin_film}) includes the shear contribution $\tau$.
Assuming the interface is not accelerated (e.g., in case of dewetting) the shear term can be neglected and the resulting equation is 
\begin{equation}\label{eq:thin_final}
    \partial_t h(\mathbf{x},t) = \nabla\cdot\left(\frac{h^3}{3\mu}\nabla p\right),
\end{equation}
which completes the task to derive the thin film equation from the Navier-Stokes equation. 
To sum up the approach in one sentence, we non-dimensionalized the variables, collected terms in orders of $\varepsilon$, used our boundary conditions and performed an integration along the vertical dimension to average out the vertical dynamics.
Of course this derivation comes short if there are additional forces to consider such as thermal fluctuations, that said in the third publication we discuss the effect of thermal fluctuations on dewetting and show how the thin film equation has to be modified.

@@

An example of a dispersion relation which follows the above equation is shown in Fig.~\ref{fig:dispertion_1}.
In this figure we can identify two relevant wave numbers, on the one hand there is $q_0$ which is the fastest growing mode indicated by a dotted line and there is $q_m = \sqrt{2}q_0$ which is the largest wave number that is not damped.
Wavemodes larger than $q_m$ are therefore damped out.
A characteristic time scale of the problem is given as
\begin{equation}
    t_0 = \frac{3\mu}{\gamma h_0^3 q_0^4},
\end{equation}
with is used to normalize the time in Fig.~\ref{fig:dispertion_1}.
Because the concept of linear stability is such an important one, we encounter it again in the second publication as well as in the third publication.
The third publication focuses only on the addition of a fluctuating term and the so called stochastic thin film equation.
The new term due to thermal fluctuations offers some interesting features on both the structure factor as well as the dispersion.

\begin{figure}
    \centering
    \includegraphics[width=0.65\textwidth]{graphics/spectratheta20.pdf}
    \caption{Dispersion relation of a spinodally dewetting thin film with and without thermal fluctuations.
    Where $S(q)$ is the structure factor depending on the wave mode $q$.
    The dots and triangles are generated using the here presented numerical method. 
    Full and dashed lines are the theoretical prediction and its fluctuating extension.
    }
    \label{fig:dispertion_1}
\end{figure}

@@

While there are many similarities between the thin film equation and the shallow water equations, there are also some major differences.
The most pressing difference between these two models is the fact that the shallow water equation is a vector equation while the thin film equation is a scalar equation.
In the shallow water equations the liquid/vapour interface is driven by hydrostatic pressure while in the thin film equation we neglect the hydrostatic pressure and introduce surface tension and capillarity.
The question that naturally arise here is: can one model be translated into the other one?
Although this question seems odd at first it is actually of great significance to show that problems can be solved with different approaches.
There are several numerical methods to iteratively solve the thin film equation.
There are as well several different numerical methods to approximate the Navier-Stokes equation.
The important contribution to our understanding of CFD is that methods build to solve the shallow water equations can actually be used to solve the thin film problem, as discussed in the second publication.
